Non-convergence in iterative imputation

### 2025-05-20 Manuscript/manuscript.qmd
plot just 1 predictor

4. Repeat over iterations ($t = 1,...,\tau$).
    a. Repeat over variables ($j = 1,...,p$).
        I. Define $\dot{\mathbf{Y}}_{-j}^t$ as the currently complete data *except* variable $Y_j$ at iteration $t$, where $\dot{\mathbf{Y}}_{-j}^t = (\dot Y_1^t,\dots,\dot Y_{j-1}^t,\dot Y_{j+1}^{t-1},\dots,\dot Y_p^{t-1})$.
        II. Draw imputation model parameters $\dot\phi_j^t$ from the posterior predictive distribution of the observed data, where $\dot\phi_j^t \sim P(\phi_j^t|Y_j^\mathrm{obs}, \dot Y_{-j}^t, R)$.
        III. Draw imputations $\dot Y_j^t$, where $\dot Y_j^t \sim P(Y_j^\mathrm{mis}|Y_j^\mathrm{obs},  \dot Y_{-j}^t, R, \dot\phi_j^t)$.
    b. End repeat $j$.
5. End repeat $t$.

@@

\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$). $B$ is defined as the variance of the collection of average $\theta$s per chain:

@@

@@

The following figures display the simulation results for the sixteen diagnostic identifier-parameter-variable pairs, and sixteen performance measure-variable pairs, contrasted to the number of iterations in the imputation algorithm. Within the figures, we split the results according to the missingness conditions [TODO: missingness mechanisms as line types, and proportion of incomplete cases as colors?]. Note that these results are averages of the $n_{sim} = 2000$ simulation repetitions.

@@

```{r fig-ac_mean}
plot_results(results, "ac_mean", 0, "MCAR")
```

@@

```{r fig-ac_sd}
plot_results(results, "ac_sd", 0, "MCAR")
```

@@

```{r fig-psrf_mean}
plot_results(results, "psrf_mean", 1, "MCAR")
```

@@

```{r fig-psrf_sd}
plot_results(results, "psrf_sd", 1, "MCAR")
```

@@

```{r fig-bias}
plot_results(results, "bias", 0, "MCAR")
```

@@

```{r fig-cov}
plot_results(results, "cov", 0.95, "MCAR")
```

@@

```{r fig-ciw}
plot_results(results, "ciw", NULL, "MCAR")
```

@@

@@

@@

>>>>

4. Repeat over iterations ($t = 1,...,\tau$).
    a. Repeat over variables ($j = 1,...,p$).

        I. Define $\dot{\mathbf{Y}}_{-j}^t$ as the currently complete data *except* variable $Y_j$ at iteration $t$, where $\dot{\mathbf{Y}}_{-j}^t = (\dot Y_1^t,\dots,\dot Y_{j-1}^t,\dot Y_{j+1}^{t-1},\dots,\dot Y_p^{t-1})$.
        II. Draw imputation model parameters $\dot\phi_j^t$ from the posterior predictive distribution of the observed data, where $\dot\phi_j^t \sim P(\phi_j^t|Y_j^\mathrm{obs}, \dot Y_{-j}^t, R)$.
        III. Draw imputations $\dot Y_j^t$, where $\dot Y_j^t \sim P(Y_j^\mathrm{mis}|Y_j^\mathrm{obs},  \dot Y_{-j}^t, R, \dot\phi_j^t)$.
    b. End repeat $j$.
5. End repeat $t$.

@@

\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$ TODO: add not between imp var cf Rubin). $B$ is defined as the variance of the collection of average $\theta$s per chain:

@@

TODO: table captions

@@

<!-- The following figures display the simulation results for non-convergence diagnostics and performance measures, contrasted to the number of iterations in the imputation algorithm. Within the figures, we split the results according to the missingness conditions. Note that these results are averages of the $n_{sim} = 2000$ simulation repetitions. -->

<!-- [TODO: missingness mechanisms as line types, and proportion of incomplete cases as colors?] -->

@@

```{r fig-ac_mean, fig.cap="Autocorrelation in the chain means. The dashed line indicates perfect stationarity, $AC=0$. The solid black line indicates the avarage autocorrelation accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "ac_mean", 0, "MCAR")
```

@@

```{r fig-ac_sd, fig.cap="Autocorrelation in the chain variances. The dashed line indicates perfect stationarity, $AC=0$. The solid black line indicates the avarage autocorrelation accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "ac_sd", 0, "MCAR")
```

@@

```{r fig-psrf_mean, fig.cap="PSRF in the chain means. The dashed line indicates perfect mixing, $\\widehat{R}=0$. The solid black line indicates the avarage PSRF accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "psrf_mean", 1, "MCAR")
```

@@

```{r fig-psrf_sd, fig.cap="PSRF in the chain variances. The dashed line indicates perfect mixing, $\\widehat{R}=0$. The solid black line indicates the avarage PSRF accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "psrf_sd", 1, "MCAR")
```

@@

```{r fig-bias, fig.height=8, fig.cap="Bias in regression coefficient $\\beta = 1$. The dashed line indicates no bias. The solid black line indicates the avarage bias accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "bias", 0, "MCAR")
```

<!-- TODO: free-y in facet grid? Y, X1 en X3 naar appendix, CI not dashed -->

@@

```{r fig-cov, fig.cap="Coverage rate of regression coefficient $\\beta = 1$. The dashed line indicates confidence-validity. The solid black line indicates the avarage coverage rate accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "cov", 0.95, "MCAR")
```

@@

```{r fig-ciw, fig.cap="Confidence interval width of regression coefficient $\\beta = 1$. The solid black line indicates the avarage CIW accross simulation repetitions, with grey lines representing the empirical confidence interval."}
plot_results2(results, "ciw", NULL, "MCAR")
```

@@

## Conclusion

@@

## Limitations

The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm. The MICE algorithm does not start in an over-dispersed state.  MICE does not rely on starting values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values. Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data). Whether this accrues to over-dispersion is questionable. Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully. The algorithm should 'escape' the initial state (based on the starting values drawn from observed data alone). Over iterations, all parameters of interest (e.g. means, variances, multivariate estimands) should converge towards a stable state. Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations. Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.

@@

## Implications
